---
title: "Bellabeat: Google Data Analytics Capstone Project"
author: "Abdul"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

The goal of this project is to analyze smart device usage data in order to gain insight into how consumers use non-Bellabeat smart devices and how to apply these insights into Bellabeat's marketing strategy.

## Ask

The following questions are to be considered:

* What are some trends in smart device usage?
* How could these trends apply to Bellabeat customers?
* How could these trends help influence Bellabeat marketing strategy?

## Prepare

* The datasets are available [here](https://www.kaggle.com/datasets/arashnic/fitbit).
* The 31 days Time frame is limited to make any solid recommendation since there are seasons involved in a given month to consider someone's health well being.

## Process

R Studio is used for the data analysis.

#### Loading installed packages for analysis

```{r Load Libraries}
library(tidyverse)
library(lubridate)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(ggpubr)
```

#### Importing and preparing datasets
```{r Assign Variables}
d_activities <- read.csv("C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Bellabeat_data/dailyActivity_merged.csv")
d_intensities <- read.csv("C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Bellabeat_data/dailyintensities_merged.csv")
h_calories <- read.csv("C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Bellabeat_data/hourlyCalories_merged.csv")
h_steps <- read.csv("C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Bellabeat_data/hourlySteps_merged.csv")
```

#### Preview Dataset
```{r Check Data}
head(d_activities)
head(d_intensities)
tail(h_calories)
tail(h_steps)
```

#### Data Cleaning
```{r Check Data type}
glimpse(d_activities)
glimpse(d_intensities)
glimpse(h_calories)
glimpse(h_steps)
```

```{r Check Duplicates}
sum(duplicated(d_activities))
sum(duplicated(d_intensities))
sum(duplicated(h_calories))
sum(duplicated(h_steps))
```

```{r Check NA values}
sum(is.na(d_activities))
sum(is.na(d_intensities))
sum(is.na(h_calories))
sum(is.na(h_steps))
```

#### Data Formatting
```{r Date Formatting}
# activities table
d_activities <- d_activities %>%
  rename(date = ActivityDate) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

# intensities table
d_intensities <- d_intensities %>%
  rename(date = ActivityDay) %>%
  mutate(date = as_date(date, format = "%m/%d/%Y"))

# h_calories table
h_calories<- h_calories %>% 
  rename(date_time = ActivityHour) %>% 
  mutate(date_time = as.POSIXct(date_time,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))

# h_steps
h_steps<- h_steps %>% 
  rename(date_time = ActivityHour) %>% 
  mutate(date_time = as.POSIXct(date_time,format ="%m/%d/%Y %I:%M:%S %p" , tz=Sys.timezone()))
```

#### Data Merging
```{r}
# Merge of activities with intensities
d_merged <- merge(d_activities, d_intensities, by = c("Id", "date"))

# Merge the calories_intensities with steps
h_merged <- merge(h_calories, h_steps, by = c("Id", "date_time"))

# Check the new table with head()
head(d_merged)
head(h_merged)
```


## Data Analysis

#### Creating Chart Tables
```{r New Tables}
Daily_avg <- d_merged %>%
    group_by(Id) %>%
    summarise(avg_steps = mean(TotalSteps), avg_distance = mean(TotalDistance), avg_calories = mean(Calories))

# Check the new table with head()
head(Daily_avg)
```

```{r New columns}
# Categorize user steps
active_users <- Daily_avg %>%
  mutate(active_users = case_when(
    avg_steps < 5000 ~ "Inactive",
    avg_steps >= 5000 & avg_steps <= 10000 ~ "Active",
    avg_steps > 10000 ~ "Very Active"))

# Check the new table with head()
head(active_users)
```

```{r}
# Table for bar charts
weekday_d <- d_merged %>%
  mutate(weekday = weekdays(date), TotalSteps, TotalDistance)

weekday_d$weekday <- ordered(weekday_d$weekday, levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
  "Friday", "Saturday", "Sunday"))

weekday_d <- weekday_d %>%
  group_by(weekday) %>%
  summarize (daily_steps = mean(TotalSteps), daily_calories = mean(Calories))

# Check the new table with head()
head(weekday_d)
```

```{r Percentage Tables}
# Create a new percentage table from active_users
users_perc <- active_users %>%
    group_by(active_users) %>%
    summarise(total = n()) %>%
    mutate(totals = sum(total)) %>%
    group_by(active_users) %>%
    summarise(total_percent = total / totals) %>%
    mutate(labels = scales::percent(total_percent))

head(users_perc)
```

## Data Visualization

```{r}
# Percentage of Active Users
users_perc %>%
  ggplot(aes(x="",y=total_percent, fill=active_users)) +
  geom_bar(stat = "identity", width = 1)+
  coord_polar("y", start=0)+
  theme_minimal()+
  theme(axis.title.x= element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(), 
        panel.grid = element_blank(), 
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size=14, face = "bold")) +
  scale_fill_manual(values = c("#85e085","#ffd480", "#ff8080", "#e6e600")) +
  geom_text(aes(label = labels),
            position = position_stack(vjust = 0.5))+
  labs(title="User's Active Based on Steps")
```

```{r}
# Steps vs Calories
d_merged %>% 
  group_by(TotalSteps, Calories) %>% 
  ggplot(aes(x = TotalSteps, y = Calories, color = Calories)) +
  geom_point() +
  geom_smooth(color = "blue") + 
  theme(legend.position = c(.8, .3),
        legend.spacing.y = unit(2, "mm"), 
        panel.border = element_rect(colour = "black", fill=NA),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black")) +
  labs(title = 'Calories vs. Total Steps',
       y = 'Calories',
       x = 'Total Steps',
       caption = 'Data Source: FitBit Fitness Tracker Data')
```


```{r}
# Bar chart for Steps
ggplot(weekday_d, aes(weekday, daily_steps)) +
      geom_col(fill = "#006699") +
      labs(title = "Daily Steps per Weekday", x= "", y = "")
```


## Key Findings

With the data cleaned and analyzed, it has provided valuable insights for Bellabeat's marketing strategy team.

* The majority of the users are moderately active and the next majority of the users are inactive (under 5,000 steps per day).
* 54% of the users didn't has insufficient amount of sleep and 42% has good sleep. It's fairly equal.
* There is a fair amount of positive correlation between steps and calories burnt per day.
* There is no correlation between the amount of steps and amount of sleep.
* Users are mostly active on Saturday, least active during Sundays, and follows up to a near-equal amount of activities on Monday and Tuesday.
* Users, on average did not get enough sleep every day. The most amount of sleep they get is only on Sunday.

## Recommendations

Keeping in mind, Bellabeat is a high-tech manufacturer of health-focused products for women and the main goal of this project is to gain insight into how consumers use non-Bellabeat smart devices and provide high-level recommendations for how these trends can inform Bellabeat's marketing strategy.
