---
title: "Cyclistic: Google Data Analytics Capstone Project"
author: "Abdul"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Cyclistic is a bike-share company that has over 5800 bicycles and over 600 docking stations. They offer various types of bikes like reclining bikes,hand tricycles and cargo bikes (which can be used by people with disabilities).

Until now, Cyclistic’s marketing strategy relied on building general awareness and appealing to broad consumer segments. One approach that helped make these things possible was the flexibility of its pricing plans: single-ride passes, full-day passes, and annual memberships. Customers who purchase single-ride or full-day passes are referred to as casual riders. Customers who purchase annual memberships are Cyclistic members. The finance analysts have concluded that annual members are much more profitable than casual riders.

## Ask

As a junior data analyst working in the marketing analyst team at Cyclistic, i am required to provide recommendations based on the following questions.

* How does casual riders and annual members use cyclistic bikes differently ?
* I am to recommend marketing strategies that will help increase the number of annual members(convert casual riders to annual members) because they are more profitable
* The recommended marketing strategies would be approved by the executive-teams

## Prepare

* The datasets are available [here](http://https//divvy-tripdata.s3.amazonaws.com/index.html).
* January 2021 - December 2021 timeline would be considered for this analysis. 
* The riders personal informations have been removed due to data privacy issues
* Each datasets contains  up to *800,000* observations and *13* columns.


## Process

R Studio is used for the data analysis.

```{r}
# Install Packages

options(repos = list(CRAN="http://cran.rstudio.com/"))
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("lubridate")
install.packages("tidyr")
install.packages("readr")
install.packages("data.table")
install.packages("janitor")
install.packages("dplyr")
```

```{r}
# Load packages

library(tidyverse)
library(ggplot2)
library(lubridate)
library(tidyr)
library(readr)
library(data.table)
library(janitor)
library(dplyr)
```

```{r}
# Import datasets

jan21<- read_csv("C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202101-divvy-tripdata.csv")
feb21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202102-divvy-tripdata.csv')
mar21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202103-divvy-tripdata.csv')
apr21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202104-divvy-tripdata.csv')
may21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202105-divvy-tripdata.csv')
jun21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202106-divvy-tripdata.csv')
jul21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202107-divvy-tripdata.csv')
aug21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202108-divvy-tripdata.csv')
sep21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202109-divvy-tripdata.csv')
oct21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202110-divvy-tripdata.csv')
nov21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202111-divvy-tripdata.csv')
dec21<- read_csv('C:/Users/Abdulkadir/Documents/HDD/Data Analysis/R_Case_studies/Cyclistic_data/202112-divvy-tripdata.csv')
```

```{r}
# Merge datasets

ride_data<- bind_rows(jan21,feb21,mar21,apr21,may21,jun21,jul21,aug21,sep21,oct21,nov21,dec21)
```

```{r}
# check merged data frame

colnames(ride_data)
head(ride_data)  
str(ride_data)  
summary(ride_data)
```

```{r}
# Remove columns not needed for analysis 

ride_data<- ride_data %>% 
  select(-c(start_station_id, end_station_id,start_lat,start_lng,end_lat,end_lng,start_station_id,end_station_id))
dim(ride_data)
```

```{r}
# Rename columns

ride_data <- ride_data %>% 
  rename(ride_type = rideable_type, 
         start_time = started_at, 
         end_time = ended_at,
         customer_type = member_casual)
```

```{r}
# Adding day, month, and year columns

ride_data <- ride_data %>% 
  mutate(year = format(as.Date(start_time), "%Y")) %>%          #year
  mutate(month = format(as.Date(start_time), "%B")) %>%         #month
  mutate(day_of_week = format(as.Date(start_time), "%A"))       #weekday
```

```{r}
# Create 'ride_lenght' column. 

ride_data <- ride_data %>% 
  mutate(ride_length = difftime(end_time, start_time)) %>%
  mutate(ride_length = as.numeric(ride_length))  # convert to numeric for                                                              calculation
  
is.numeric(ride_data$ride_length) # to check

  head(ride_data)
```

```{r Removing unwanted data}

# Checked 'start_station_name' for rows when the bikes was taken for quality test

nrow(subset(ride_data, start_station_name %like% "TEST"))
nrow(subset(ride_data, start_station_name %like% "Test"))
nrow(subset(ride_data, start_station_name %like% "test"))

# Deleting the rows

ride_data_1<- ride_data[!(ride_data$ride_length<1 |
                          ride_data$start_station_name %like% "Test" ),]
dim(ride_data_1)

# Removing rows with nulls

ride_data_clean <- drop_na(ride_data_1)
dim(ride_data_clean)
```


## Analyze

```{r}
#Checking the count of each customer type

table(ride_data_clean$customer_type)

#aggregating total ride length by customer type

setNames(aggregate(ride_length ~ customer_type, ride_data_clean,sum),
         c('customer_type', 'total_ride_length(min)'))
```

```{r}
# Comparison between customer types based on ride length 

ride_data_clean %>% 
  group_by(customer_type) %>% 
  summarize(average_ride_length = mean(ride_length),
            median_ride_length = median(ride_length),
            max_ride_length = max(ride_length),
            min_ride_length = min(ride_length))
```

```{r}
# members vs casual riders difference depending on total rides taken

ride_data_clean %>% 
    group_by(customer_type) %>% 
    summarise(ride_count = length(ride_id), ride_percentage = (length(ride_id) / nrow(ride_data_clean)) * 100)
```

```{r}
# Average ride time by each day for each customer type
 
ride_data_clean$day_of_week <- ordered(ride_data_clean$day_of_week,
          levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
  
ride_data_clean %>% 
  group_by(customer_type, day_of_week) %>%  
  summarise(number_of_rides = n(),  
  average_ride_length = mean(ride_length),.groups="drop") %>%
  arrange(customer_type, day_of_week)
```


## Visualize

```{r}
# Customer Distribution

ggplot(ride_data_clean, aes(x = customer_type, fill = customer_type)) +
    geom_bar() +
    labs(x = "Members vs Casuals", y= "Number Of Rides", title = "Members vs Casuals distribution")
```

```{r}

ride_data_clean %>%
group_by(customer_type, day_of_week) %>%  
  summarise(number_of_rides = n(),  
  average_ride_length = mean(ride_length),.groups="drop") %>%
  arrange(customer_type, day_of_week) %>%
ggplot(aes(x = day_of_week, y = average_ride_length, fill = customer_type)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average ride time by customer type Vs. Day of the week")
```

```{r}

ride_data_clean %>%
group_by(customer_type, month) %>%  
  summarise(number_of_rides = n(),  
  average_ride_length = mean(ride_length),.groups="drop") %>%
  arrange(customer_type, month) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = customer_type)) +
  labs(title ="Total rides by customer type Vs. Month", x = "Month", y= "Number Of Rides") +
  theme(axis.text.x = element_text(angle = 45)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

```{r}

ride_data_clean %>%
    ggplot(aes(ride_type, fill= customer_type)) +
    labs(x="Hour of the day", title="Bike demand by hour in a day") +
    geom_bar()
```

```{r}
ride_data_clean %>%
    ggplot(aes(ride_type, fill = customer_type)) +
    geom_bar() +
    labs(x="Hour of the day", title="Bike demand per hour by day of the week") +
    facet_wrap(~ day_of_week)
```

## Share

##### Differences between members and casuals:

* Members have a higher preference for classic bikes.
* Casuals riders have the most data points on weekends.
* Casuals riders have more ride duration than members.
* Members are more active between 7am and 10am while casuals between 3pm and 12am.
* Members have a more fixed routine while casual rider are mostly active on weekends.

## Act

* Offer a weekend-only membership at a different price point than the full annual membership.
* Create marketing campaigns which can be sent via email, or advertisement in the docking stations explaining why annual member is beneficial. Campaigns should be placed at the peak months of the year.
